---
title: "Dose escalation dataset example - population level cellular barcoding analysis"
output: html_notebook
---

## import libraries
```{r}
library(bartools)
library(tidyverse)
```

# setup data output directories
```{r}
plot.dir <- "./plots"
results.dir <- "./results"
dir.create(plot.dir, showWarnings = F)
dir.create(results.dir, showWarnings = F)
```

# Data import and QC

## import dataset
Here we examined the clonal dynamics of Acute Myeloid Leukaemia (AML) cells in response to escalating doses of Cytarabine (AraC) a conventional chemotherapy used routinely in the clinic, or IBET-151, a targeted therapy against the BET bromodomain family of transcriptional co-activators which has shown pre-clinical efficacy against MLL-rearranged AML.

Mouse MLL-AF9 fusion oncogene containing AML cells were barcoded with the SPLINTR cellular barcoding system and cultured in vitro with increasing doses of IBET-151, AraC or a corresponding vehicle control (DMSO). At each dose escalation, cells were sampled for bulk barcode sequencing. The dataset (hereafter the “dose escalation” dataset) comprises bulk barcode-seq data per timepoint. 

BARtab was previously run on all 86 datasets and the counts are imported below.

NB: sample labels for `IBET-gradual-TR1-TP5` and `ARAC-gradual-TR1-TP5` were switched in sequencing and are fixed in the counts table and samplesheet manually before importing into this analysis

```{r}
sampletable <- read.csv("../../datasets/dose_escalation_02/dose_escalation_02_samplesheet.csv", header = T)

# reformat names and order of counts table to match samplesheet
counts <- read.delim("../../datasets/dose_escalation_02/dose_escalation_02_counts.tsv", header = T, row.names = 1)
counts <- counts[,order(colnames(counts))]
colnames(counts) <- gsub("_S[0-9][0-9]_R[1]_001", "", colnames(counts))
colnames(counts) <- gsub("_S[0-9]_R[1]_001", "", colnames(counts))
colnames(counts) <- gsub("\\.", "-", colnames(counts))

# check ordering is correct before import
sampletable$Sample == colnames(counts)

# import count data and metadata into a DGEList object
dose <- DGEList(counts = counts, samples = sampletable, group = sampletable$Run, remove.zeros = T)
dim(dose)

# reformat sample names
colnames(dose) <- paste(dose$samples$Run, dose$samples$PCR_Replicate, sep = "_")

```

## Remove rows with no data
```{r}
keeprows = rowSums(dose$counts) >= 1
dose.filtered <- dose[keeprows,]
dose.filtered.counts <- dose.filtered$counts
dim(dose.filtered.counts)
```

## Examine BARtab output for dataset
```{r, eval = FALSE}
# this will not work if you have not preprocessed the raw data with BARtab
bartab.dir <- "../../datasets/dose_escalation_02/preprocessed_data/"

# filtered reads
plotBARtabFilterQC(bartab.dir)

# mapped reads
plotBARtabMapQC(bartab.dir)
```

## Examine total read counts across samples
```{r}
plotReadCounts(dose.filtered$counts, group = dose.filtered$samples$Treatment)
```

Some samples have very low total counts due to very poor mappability and are unlikely to be informative. We will remove these from the analysis. 

As standard we apply a 5th percentile threshold on the data, removing any samples with total counts below this threshold. 

## Filter out samples below 5th percentile threshold for total read counts
```{r}
threshold <- quantile(x = colSums(dose.filtered$counts), .05)
# 5% = 57473.5

samps.to.remove <- which(colSums(dose.filtered$counts) < threshold)
samps.to.remove

dose.filtered <- dose.filtered[,-samps.to.remove]
dim(dose.filtered)
```

## Filter samples using absolute or relative abundance thresholds. 
```{r}
thresholdCounts(dose.filtered, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Treatment") + coord_flip()
ggsave(file.path(plot.dir,"thresholdBarcodes_Treatment.pdf"), width = 7, height = 10)

thresholdCounts(dose.filtered, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Strategy") + coord_flip()
ggsave(file.path(plot.dir,"thresholdBarcodes_Strategy.pdf"), width = 7, height = 10)

thresholdCounts(dose.filtered, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Dose") + coord_flip()
ggsave(file.path(plot.dir,"thresholdBarcodes_Dose.pdf"), width = 7, height = 10)

thresholdCounts(dose.filtered, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Timepoint") + coord_flip()
ggsave(file.path(plot.dir,"thresholdBarcodes_Timepoint.pdf"), width = 7, height = 10)
```

## filter samples with threshold 10 reads across at least 4 samples
```{r}
dose.filtered <- thresholdCounts(dose.filtered, type = "absolute", threshold = 10, min.samps = 4, plot = F)
dim(dose.filtered)
```

## Normalise each sample to counts per million
```{r}
dose.cpmnorm <- as.data.frame(cpm(dose.filtered$counts, normalized.lib.sizes = T))
```

## divide samples into treatments
```{r}
ibet <- dose.filtered[,which(dose.filtered$samples$Treatment == "IBET" | dose.filtered$samples$Treatment == "DMSO"| dose.filtered$samples$Treatment == "TP0")]

arac <- dose.filtered[,which(dose.filtered$samples$Treatment == "ARAC" | dose.filtered$samples$Treatment == "DMSO"| dose.filtered$samples$Treatment == "TP0")]
```

## plot raw counts per sample
```{r}
plotReadCounts(dose.filtered$counts, group = dose.filtered$samples$Treatment) + 
  geom_vline(xintercept = threshold)
ggsave(file.path(plot.dir,"rawcounts_Treatment.pdf"), width = 7, height = 10)

plotReadCounts(ibet$counts, group = ibet$samples$Treatment) + 
  geom_vline(xintercept = threshold)
ggsave(file.path(plot.dir,"rawcounts_ibet_Treatment.pdf"), width = 7, height = 10)

plotReadCounts(arac$counts, group = arac$samples$Treatment) + 
  geom_vline(xintercept = threshold)
ggsave(file.path(plot.dir,"rawcounts_arac_Treatment.pdf"), width = 7, height = 10)
```

## plot total counts for detected barcodes
```{r}
pdf("./plots/Total_counts.pdf")
plotBarcodeCounts(dose.filtered$counts, log10 = F)
plotBarcodeCounts(dose.filtered$counts, log10 = T)

# ibet
plotBarcodeCounts(ibet$counts, log10 = F)
plotBarcodeCounts(ibet$counts, log10 = T)

# arac
plotBarcodeCounts(arac$counts, log10 = F)
plotBarcodeCounts(arac$counts, log10 = T)
dev.off()
```
## calculate correlation between PCR replicates
```{r}
corrs <- calcReplicateCorr(dose.filtered, group = "group", corr.thresh = 0.9, return = "all")
which(corrs < 0.8)

```

## plot correlation between PCR replicates
```{r}
samps <- unique(dose.filtered$samples$Run)
# remove unpaired sample
samps <- samps[-29]

pdf("./plots/Replicate_corr.pdf", width = 6, height = 6)
lapply(samps, function(x){
  df <- dose.filtered[,dose.filtered$samples$group %in% as.character(x)]
  plotBarcodeRegression(df, samp1 = colnames(df)[[1]], samp2 = colnames(df)[[2]], rug = T, trans = "log10")
})
dev.off()
```

## example good replicate correlation
```{r}
df <- dose.filtered[,dose.filtered$samples$group == "TP0"]
plotBarcodeRegression(df, samp1 = colnames(df)[[1]], samp2 = colnames(df)[[2]], rug = T, trans = "log1p")
```

## example poor replicate correlation
```{r}
df <- dose.filtered[,dose.filtered$samples$group == "ARAC_high_dose_TR1_TP1"]
plotBarcodeRegression(df, samp1 = colnames(df)[[1]], samp2 = colnames(df)[[2]], rug = T, trans = "log1p")
```

## plot all pairwise correlations
```{r}
plotBarcodeCorrelation(dose.filtered$counts, upper = F, clustered = T)
ggsave("./plots/pearson_corr_all_PCR_reps_clustered.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = F, clustered = F)
ggsave("./plots/pearson_corr_all_PCR_reps_not-clustered.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = T, clustered = T)
ggsave("./plots/pearson_corr_all_PCR_reps_clustered_upper.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = T, clustered = F)
ggsave("./plots/pearson_corr_all_PCR_reps_not-clustered_upper.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = F, clustered = T)
ggsave("./plots/spearman_corr_all_PCR_reps_clustered.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = F, clustered = F)
ggsave("./plots/spearman_corr_all_PCR_reps_not-clustered.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = T, clustered = T)
ggsave("./plots/spearman_corr_all_PCR_reps_clustered_upper.pdf", width = 10, height = 9)

plotBarcodeCorrelation(dose.filtered$counts, upper = T, clustered = F)
ggsave("./plots/spearman_corr_all_PCR_reps_not-clustered_upper.pdf", width = 10, height = 9)
```

## Collapse PCR replicates in object
```{r}
# take the average of PCR technical duplicates within a sample
dose.filtered.collapsed <- collapseReplicates(dose.filtered, groupby = dose.filtered$samples$group, by = "mean")
```

The result is a clean barcode sequencing dataset ready for further investigation and visualisation.
```{r}
dim(dose.filtered.collapsed)
```

Spurious barcodes can appear in endpoint samples but not in the baseline sample from which they were derived. At the experimental level a choice can be made to limit analysis to barcodes detected at this baseline timepoint which can further clean up downstream analyses

## filter barcodes not present in T0
```{r}
remove <- which(dose.filtered.collapsed$counts[,"TP0"] == 0)
dose.filtered.collapsed <- dose.filtered.collapsed[-remove,]
```

```{r}
dim(dose.filtered.collapsed)
```

## save preprocesed dataset
```{r}
saveRDS(dose.filtered.collapsed, file = "./dose2.rds")
```

# read in preprocessed dataset and subset into individual treatments
```{r}
dge <- readRDS("dose2.rds")

# simplify barcode labels
rownames(dge) <- gsub("BFP_Barcode", "BC", rownames(dge))
```


## generate ibet and AraC specific datasets
```{r}
ibet <- dge[,which(dge$samples$Treatment == "IBET" | dge$samples$Treatment == "DMSO"| dge$samples$Treatment == "TP0")]

arac <- dge[,which(dge$samples$Treatment == "ARAC" | dge$samples$Treatment == "DMSO"| dge$samples$Treatment == "TP0")]
```

# split dataset into replicates
```{r}
dge.TR1 <- dge[,which(dge$samples$Technical_Replicate == "TR1")]
dge.TR2 <- dge[,which(dge$samples$Technical_Replicate == "TR2")]
```

# confirm count thresholds
```{r}
thresholdCounts(dge, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Treatment") + coord_flip()
ggsave("./plots/thresholdBarcodes_collapsed_Treatment.pdf", width = 6, height = 8)

thresholdCounts(dge, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Dose") + coord_flip()
ggsave("./plots/thresholdBarcodes_collapsed_Dose.pdf", width = 6, height = 8)

thresholdCounts(dge, type = "absolute", threshold = 10, min.samps = 4, plot = T, group = "Timepoint") + coord_flip()
ggsave("./plots/thresholdBarcodes_collapsed_Timepoint.pdf", width = 6, height = 8)
```

# Visualisation

bartools includes a range of visualisation options for examining barcode-seq datasets. 

## Bubble plot
```{r}
plotBarcodeBubble(dge$counts, proportion.cutoff = 10, labels = T)
ggsave("./plots/Bubbleplot_all_collapsed.pdf")

plotBarcodeBubble(ibet$counts, proportion.cutoff = 10, labels = T)
ggsave("./plots/Bubbleplot_ibet_collapsed.pdf")

plotBarcodeBubble(arac$counts, proportion.cutoff = 10, labels = T)
ggsave("./plots/Bubbleplot_arac_collapsed.pdf")
```


## Ordered Bubble plots
```{r}
plotOrderedBubble(dge$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed.pdf")

plotOrderedBubble(dge$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = dge$samples, group = "Treatment", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_Treatment.pdf")

plotOrderedBubble(dge$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = dge$samples, group = "Strategy", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_Strategy.pdf")

plotOrderedBubble(dge$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = dge$samples, group = "Timepoint", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_Timepoint.pdf")
```

### IBET only
```{r}
plotOrderedBubble(ibet$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_ibet.pdf")

plotOrderedBubble(ibet$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = ibet$samples, group = "Treatment", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_ibet_Treatment.pdf")

plotOrderedBubble(ibet$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = ibet$samples, group = "Strategy", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_ibet_Strategy.pdf")

plotOrderedBubble(ibet$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = ibet$samples, group = "Timepoint", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_ibet_Timepoint.pdf")
```

### AraC
```{r}
plotOrderedBubble(arac$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_arac.pdf")

plotOrderedBubble(arac$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = arac$samples, group = "Treatment", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_arac_Treatment.pdf")

plotOrderedBubble(arac$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = arac$samples, group = "Strategy", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_arac_Strategy.pdf")

plotOrderedBubble(arac$counts, proportion.cutoff = 10, labels = T, orderSample = "TP0", filterLow = T, samples = arac$samples, group = "Timepoint", filter.cutoff = .01)
ggsave("./plots/orderedBubble_collapsed_arac_Timepoint.pdf")

```

## Barcode Plot
```{r}
pdf("./plots/plot_barcode_histograms.pdf")
plotBarcodeHistogram(dge$counts, sample = dge$samples$group[[40]], top = 50, name = "All samples")
plotBarcodeHistogram(ibet$counts, sample = ibet$samples$group[[28]], top = 50, name = "IBET")
plotBarcodeHistogram(arac$counts, sample = arac$samples$group[[21]], top = 50, name = "AraC")
dev.off()

plotBarcodeHistogram(dge$counts, sample = dge$samples$group[[40]], top = 50, name = "All samples")
plotBarcodeHistogram(ibet$counts, sample = ibet$samples$group[[28]], top = 50, name = "IBET")
plotBarcodeHistogram(arac$counts, sample = arac$samples$group[[21]], top = 50, name = "AraC")

```

## Timeseries Plot
```{r}
plotBarcodeTimeseries(dge$counts, seed = 101, top = 50)
ggsave("./plots/timeseries_all_top50.pdf")
plotBarcodeTimeseries(ibet$counts, seed = 101, top = 50)
ggsave("./plots/timeseries_ibet_top50.pdf")
plotBarcodeTimeseries(arac$counts, seed = 101, top = 50)
ggsave("./plots/timeseries_arac_top50.pdf")

# replicate level
plotBarcodeTimeseries(dge.TR1$counts, seed = 101, top = 50)
ggsave("./plots/timeseries_dge.TR1_top50.pdf")
plotBarcodeTimeseries(dge.TR2$counts, seed = 101, top = 50)
ggsave("./plots/timeseries_dge.TR2_top50.pdf")
```

### merge Treatment & Strategy groups and facet manually
```{r}
dge.TR1$samples$TreatStrat <- factor(paste(dge.TR1$samples$Treatment, dge.TR1$samples$Strategy, sep = "_"), levels = c("TP0_TP0","DMSO_DMSO","ARAC_gradual","ARAC_high_dose","IBET_gradual","IBET_high_dose" ))
dge.TR1$samples$group <- factor(dge.TR1$samples$group, levels = rev(levels(dge.TR1$samples$group)))

plotBarcodeTimeseries(dge.TR1$counts[, dge.TR1$samples$Strategy %in% c("TP0", "DMSO", "gradual")], seed = 101, top = 50)
plotBarcodeTimeseries(dge.TR1$counts[, dge.TR1$samples$TreatStrat == "DMSO_DMSO"| dge.TR1$samples$TreatStrat == "TP0_TP0"], seed = 101, top = 50)
plotBarcodeTimeseries(dge.TR1$counts[, dge.TR1$samples$TreatStrat == "IBET_gradual"| dge.TR1$samples$TreatStrat == "TP0_TP0"], seed = 101, top = 50)
```

## Principal Components Analysis
A global level PCA analysis is a good way to get a high level understanding of the similarities and differences between samples. 
```{r}
plotBarcodePCA(dge, intgroup = c("Treatment", "Timepoint")) + scale_color_manual(values = c(viridis::viridis(6)[1:5], rep("grey80", 5), viridis::inferno(6)[1:5], "red"))
ggsave("./plots/PCA_all_treatment_timepoint.pdf", width = 5, height = 4)
plotBarcodePCA(dge, intgroup = c("Treatment", "Dose"))
ggsave("./plots/PCA_all_treatment_dose.pdf", width = 5, height = 4)

plotBarcodePCA(dge, intgroup = "Treatment")
ggsave("./plots/PCA_all_treatment.pdf", width = 5, height = 4)
plotBarcodePCA(ibet, intgroup = "Dose")
ggsave("./plots/PCA_ibet_dose.pdf", width = 5, height = 4)
plotBarcodePCA(arac, intgroup = "Dose")
ggsave("./plots/PCA_arac_dose.pdf", width = 5, height = 4)
```

## Heatmaps
```{r}
plotBarcodeHeatmap(dge$counts, N = 10, samples = dge$samples, group = c("Treatment", "Timepoint", "Strategy"))
plotBarcodeHeatmap(ibet$counts, N = 10, samples = ibet$samples, group = c("Treatment", "Timepoint", "Strategy"))
plotBarcodeHeatmap(arac$counts, N = 10, samples = arac$samples, group = c("Treatment", "Timepoint", "Strategy"))

plotBarcodeHeatmap(dge.TR1$counts, N = 10, samples = dge.TR1$samples, group = c("Treatment", "Timepoint", "Strategy"))
plotBarcodeHeatmap(dge.TR2$counts, N = 10, samples = dge.TR2$samples, group = c("Treatment", "Timepoint", "Strategy"))
```

### write heatmaps
```{r}
pdf("./plots/heatmaps_top10.pdf", width = 5, height = 8)
plotBarcodeHeatmap(dge$counts, N = 10, samples = dge$samples, group = c("Treatment", "Timepoint", "Strategy"))
plotBarcodeHeatmap(ibet$counts, N = 10, samples = ibet$samples, group = c("Treatment", "Timepoint", "Strategy"))
plotBarcodeHeatmap(arac$counts, N = 10, samples = arac$samples, group = c("Treatment", "Timepoint", "Strategy"))
dev.off()
```

## Diversity metrics
```{r}
diversity <- calcDivIndexes(dge$counts, group = dge$samples$Treatment)
diversity
write.csv(diversity, file = "./diversity_calculations.csv")
```

### plot diversity metrics
```{r}
plotDivIndexes(div = diversity, metric = "shannon", type = "point", group = dge$samples$Treatment)
plotDivIndexes(div = diversity, metric = "shannon", type = "point", group = dge$samples$Treatment) + facet_wrap(~diversity$group)
plotDivIndexes(div = diversity, metric = "shannon", group = dge$samples$Treatment)
ggsave("./plots/shannon_histogram.pdf")
```

### TP5 samples
```{r}
TP5.samps <- diversity[grep("high_dose", diversity$name, perl = T, invert = T),]
TP5.samps <- TP5.samps[grep("TP5|TP0", TP5.samps$name, perl = T, invert = F),]

plotDivIndexes(div = TP5.samps, metric = "shannon", group = TP5.samps$group)
ggsave("./plots/shannon_histogram_TP5.pdf")
```

### gradual dose escalation samples
```{r}
gradual.samps <- diversity[grep("high_dose", diversity$name, perl = T, invert = T),]
TR1.samps <- gradual.samps[grep("TR1|TP0", gradual.samps$name, perl = T, invert = F),]
TR2.samps <- gradual.samps[grep("TR2|TP0", gradual.samps$name, perl = T, invert = F),]
plotDivIndexes(div = TR1.samps, metric = "shannon", group = TR1.samps$group)
ggsave("./plots/shannon_histogram_TR1.pdf")
plotDivIndexes(div = TR2.samps, metric = "shannon", group = TR2.samps$group)
ggsave("./plots/shannon_histogram_TR2.pdf")
```

### Plot distance between samples
```{r}
plotBarcodeDistance(dge$counts, method = "euclidean", clustered = F)
ggsave("./plots/euclidean_dist_all.pdf")
plotBarcodeDistance(ibet$counts, method = "euclidean", clustered = F)
ggsave("./plots/euclidean_dist_ibet.pdf")
plotBarcodeDistance(arac$counts, method = "euclidean", clustered = F)
ggsave("./plots/euclidean_dist_arac.pdf")

# replicate level 
plotBarcodeDistance(dge.TR1$counts, method = "euclidean", clustered = F)
ggsave("./plots/euclidean_dist_TR1.pdf")
plotBarcodeDistance(dge.TR2$counts, method = "euclidean", clustered = F)
ggsave("./plots/euclidean_dist_TR2.pdf")



```

## Plot correlation between samples
```{r}
plotBarcodeCorrelation(cpm(dge), method = "pearson", clustered = T)
ggsave("./plots/pearson_collapsed_all.pdf")
plotBarcodeCorrelation(cpm(ibet$counts), method = "pearson", clustered = T)
ggsave("./plots/pearson_collapsed_ibet.pdf")
plotBarcodeCorrelation(cpm(arac$counts), method = "pearson", clustered = T)
ggsave("./plots/pearson_collapsed_arac.pdf")

plotBarcodeCorrelation(cpm(dge), method = "spearman", clustered = T)
ggsave("./plots/spearman_collapsed_all.pdf")
plotBarcodeCorrelation(cpm(ibet$counts), method = "spearman", clustered = T)
ggsave("./plots/spearman_collapsed_ibet.pdf")
plotBarcodeCorrelation(cpm(arac$counts), method = "spearman", clustered = T)
ggsave("./plots/spearman_collapsed_arac.pdf")

# replicate level
plotBarcodeCorrelation(cpm(dge.TR1), method = "spearman", clustered = T)
ggsave("./plots/spearman_collapsed_TR1.pdf")
plotBarcodeCorrelation(cpm(dge.TR2), method = "spearman", clustered = T)
ggsave("./plots/spearman_collapsed_TR2.pdf")

```

## Calculate Percentile barcodes
```{r}
percentile.bc <- calcPercentileBarcodes(dge, percentile = 0.95)
```

## plot detected barcodes 
```{r}
plotDetectedBarcodes(dge, group = "Treatment", percentile = 0.95) + coord_flip()
ggsave("./plots/detected_barcodes_all_95th.pdf", width = 7, height = 10)

plotDetectedBarcodes(ibet, group = "Treatment", percentile = 0.95) + coord_flip()
ggsave("./plots/detected_barcodes_ibet_95th.pdf", width = 7, height = 6)

plotDetectedBarcodes(arac, group = "Treatment", percentile = 0.95) + coord_flip()
ggsave("./plots/detected_barcodes_arac_95th.pdf", width = 7, height = 6)
```

## compare dominant barcodes across treatments
### calculate dominant barcodes above 1pct thresh
```{r}
library(VennDetail)

# ibet dominant barcodes per treatment strategy
ibet.bc <- getDominantBarcodes(dge[,dge$samples$Treatment == "IBET"], pct.thresh = 1)
ibet.bc.gradual <- unique(unlist(ibet.bc[1:10]))
ibet.bc.high <- unique(unlist(ibet.bc[11:18]))

ibet.res <- venndetail(list(ibet.gradual = ibet.bc.gradual, ibet.high = ibet.bc.high))
plot(ibet.res)

# arac dominant barcodes per treatment strategy
arac.bc <- getDominantBarcodes(dge[,dge$samples$Treatment == "ARAC"], pct.thresh = 1)
arac.bc.gradual <- unique(unlist(arac.bc[1:8]))
arac.bc.high <- unique(unlist(arac.bc[9:12]))

arac.res <- venndetail(list(arac.gradual = arac.bc.gradual, arac.high = arac.bc.high))
plot(arac.res, show.number = T)

pdf("./plots/all_dom_bc_venn.pdf", width = 5, height = 5)
plot(venndetail(list(arac.gradual = arac.bc.gradual, arac.high = arac.bc.high,ibet.gradual = ibet.bc.gradual, ibet.high = ibet.bc.high)), show.number = T)
dev.off()
```

### upset plot
```{r}
library(UpSetR)
bc.dat <- list(arac.gradual = arac.bc.gradual, 
               arac.high = arac.bc.high,
               ibet.gradual = ibet.bc.gradual, 
               ibet.high = ibet.bc.high)
upset(fromList(bc.dat), 
      sets = c("ibet.gradual", "ibet.high", "arac.gradual", "arac.high"), 
      nintersects = 8,
      )
```

## Clone level analysis
```{r}
#ibet
plotBarcodeBoxplot(dge, barcodes = c("BC_218246", "BC_110908"), group = "Treatment", trans = "log10", conditions =  c("ARAC", "IBET", "DMSO"), point = T, violin = T) + scale_fill_manual(values = rev(ggpubr::get_palette("npg", 4)[2:4]))
ggsave("./plots/barcode_boxplots_ibet.pdf", width = 5, height = 3)

# arac
plotBarcodeBoxplot(dge, barcodes = c("BC_525057", "BC_250687"), group = "Treatment", trans = "log10", conditions =  c("ARAC", "IBET", "DMSO"), point = T, violin = T) + scale_fill_manual(values = rev(ggpubr::get_palette("npg", 4)[2:4]))
ggsave("./plots/barcode_boxplots_arac.pdf", width = 5, height = 3)

# dmso
plotBarcodeBoxplot(dge, barcodes = c("BC_341397", "BC_250687"), group = "Treatment", trans = "log10", conditions =  c("ARAC", "IBET", "DMSO"), point = T, violin = T) + scale_fill_manual(values = rev(ggpubr::get_palette("npg", 4)[2:4]))
ggsave("./plots/barcode_boxplots_dmso.pdf", width = 5, height = 3)

```

## Comparing abundance
We can statistically test for barcodes / tags that are over / underrepresented in a group of samples relative to another using the internal edgeR framework. barista contains a convenience wrapper for this functionality
```{r}
ibet.abundance <- compareAbundance(ibet, meta = "Strategy", condition1 = "gradual", "high_dose", logFC.cutoff = 3, pval.cutoff = 0.01)
ibet.abundance
```
```{r}
sessionInfo()
```

