---
title: "Single-cell comparison BARtab FateMap"
output: html_notebook
---

This notebook compares the results of barcode extraction from experiment FM02 from the original FateMap paper (Goyal et al. 2023). 

```{r}
library(tidyverse)
library(ggvenn)
```

## Data loading and preprocessing

```{r}
results_dir <- "/dawson_genomics/Projects/bartools_bartab_paper/results/sc_yogesh_comparison/"
plots_dir <- "/dawson_genomics/Projects/bartools_bartab_paper/results/sc_yogesh_comparison/plots/"
```

### Load whitelists

Read in barcode whitelists to compare how many cells were annotated with barcodes. 
Remove -1 ending of barcodes. 
```{r}
input1Directory <-
  "/dawson_genomics/Projects/bartools_bartab_paper/yogesh_data/sc_data/"

whitelist_s1 <- gsub("..$", "",
                     read.table(
                       paste0(input1Directory, "GSM7434409_FM02_A_1uMPLX_barcodes.tsv.gz"),
                       stringsAsFactors = F
                     )$V1)

whitelist_s2 <- gsub("..$", "",
                     read.table(
                       paste0(input1Directory, "GSM7434410_FM02_B_100nMPLX_barcodes.tsv.gz"),
                       stringsAsFactors = F
                     )$V1)

whitelist_s3 <- gsub("..$", "",
                     read.table(
                       paste0(input1Directory, "GSM7434411_FM02_C_5nMT_barcodes.tsv.gz"),
                       stringsAsFactors = F
                     )$V1)

whitelist_s4 <- gsub("..$", "",
                     read.table(
                       paste0(input1Directory, "GSM7434412_FM02_D_5nMT100nMP_barcodes.tsv.gz"),
                       stringsAsFactors = F
                     )$V1)
```

### Load and preprocess FateMap results

Read in FateMap results for FM02, remove unnecessary columns and uniq rows. 
Code adapted from `https://github.com/arjunrajlaboratory/FateMap_Goyal2023/blob/main/plotScripts/FM01_to_FM05/FM02_sample1Sample2_V2_hg19.R`.

Published preprocessed data contained cell barcodes not present in the whitelists. 
Therefor, we re-ran FateMap with the provided code, using the published whitelists. 

```{r}
# home2Directory <- "/dawson_genomics/Projects/bartools_bartab_paper/yogesh_data/sc_data/Processed_FM01_FM03_FM04_FM05_FM06_FM08_FM09_FM10/FM02/"
home2Directory <- "/dawson_genomics/Projects/bartools_bartab_paper/results/sc_yogesh_comparison/fatemap_results_fm02/stepThree/"
barcode50 = as_tibble(read.table(
  paste0(home2Directory, 'stepThreeStarcodeShavedReads.txt'),
  stringsAsFactors = F,
  header = F
))
barcode50 = barcode50 %>%
  dplyr::rename(
    cellID = V1,
    UMI = V2,
    BC50StarcodeD8 = V4,
    sampleNum = V8
  ) %>%
  select(-c(V3, V5, V6, V7)) %>%
  unique() %>%
  select(-UMI)
```

Count UMIs per barcode per cell. 

```{r}
barcode50_counts <- barcode50 %>%
  group_by(cellID, BC50StarcodeD8, sampleNum) %>%
  summarise(nUMI = length(sampleNum))
```

Save intermediate results, because takes a lot of memory (>12GB) to load the data.  
`barcode50_counts` is only 300MB

```{r}
write.csv(barcode50_counts, file.path(results_dir, "fm02_barcode50_counts.csv"))
```

```{r}
barcode50_counts <- read.csv(file.path(results_dir, "fm02_barcode50_counts.csv"), row.names = 1)
```

Clear memory. 

```{r}
rm(barcode50)
```

Split into separate samples. 

```{r}
timemachine_s1 <- barcode50_counts %>%
  filter(sampleNum == 1) %>%
  select(-sampleNum) %>%
  rename(bc.umi.count = nUMI, cellid = cellID, barcode = BC50StarcodeD8)

timemachine_s2 <- barcode50_counts %>%
  filter(sampleNum == 2) %>%
  select(-sampleNum) %>%
  rename(bc.umi.count = nUMI, cellid = cellID, barcode = BC50StarcodeD8)

timemachine_s3 <- barcode50_counts %>%
  filter(sampleNum == 3) %>%
  select(-sampleNum) %>%
  rename(bc.umi.count = nUMI, cellid = cellID, barcode = BC50StarcodeD8)

timemachine_s4 <- barcode50_counts %>%
  filter(sampleNum == 4) %>%
  select(-sampleNum) %>%
  rename(bc.umi.count = nUMI, cellid = cellID, barcode = BC50StarcodeD8)
```

Clear memory. 

```{r}
rm(barcode50_counts)
```

Number of cells per sample that were not present in the published whitelists. 

```{r}
sum(!unique(timemachine_s1$cellid) %in% whitelist_s1)
sum(!unique(timemachine_s2$cellid) %in% whitelist_s2)
sum(!unique(timemachine_s3$cellid) %in% whitelist_s3)
sum(!unique(timemachine_s4$cellid) %in% whitelist_s4)
```
For published FateMap results: 
[1] 61
[1] 181
[1] 31
[1] 15

Remove cells from timemachine output that are not in the whitelist. 
Not necessary with the re-analyzed data. 

```{r}
# timemachine_s1 <- timemachine_s1[timemachine_s1$cellid %in% whitelist_s1,]
# timemachine_s2 <- timemachine_s2[timemachine_s2$cellid %in% whitelist_s2,]
# timemachine_s3 <- timemachine_s3[timemachine_s3$cellid %in% whitelist_s3,]
# timemachine_s4 <- timemachine_s4[timemachine_s4$cellid %in% whitelist_s4,]
```

QC plots for FateMap. 

```{r}
bartools::plotBarcodesPerCell(timemachine_s1)
ggsave(file.path(plots_dir, "FateMap_barcodes_per_cell_FM02_S1.pdf"))
bartools::plotUmiPerBarcode(timemachine_s1)
ggsave(file.path(plots_dir, "FateMap_umi_per_barcode_FM02_S1.pdf"))
bartools::plotUmiFilterThresholds(timemachine_s1)
ggsave(file.path(plots_dir, "FateMap_umi_filter_thresholds_FM02_S1.pdf"))

bartools::plotBarcodesPerCell(timemachine_s2)
ggsave(file.path(plots_dir, "FateMap_barcodes_per_cell_FM02_S2.pdf"))
bartools::plotUmiPerBarcode(timemachine_s2)
ggsave(file.path(plots_dir, "FateMap_umi_per_barcode_FM02_S2.pdf"))
bartools::plotUmiFilterThresholds(timemachine_s2)
ggsave(file.path(plots_dir, "FateMap_umi_filter_thresholds_FM02_S2.pdf"))

bartools::plotBarcodesPerCell(timemachine_s3)
ggsave(file.path(plots_dir, "FateMap_barcodes_per_cell_FM02_S3.pdf"))
bartools::plotUmiPerBarcode(timemachine_s3)
ggsave(file.path(plots_dir, "FateMap_umi_per_barcode_FM02_S3.pdf"))
bartools::plotUmiFilterThresholds(timemachine_s3)
ggsave(file.path(plots_dir, "FateMap_umi_filter_thresholds_FM02_S3.pdf"))

bartools::plotBarcodesPerCell(timemachine_s4)
ggsave(file.path(plots_dir, "FateMap_barcodes_per_cell_FM02_S4.pdf"))
bartools::plotUmiPerBarcode(timemachine_s4)
ggsave(file.path(plots_dir, "FateMap_umi_per_barcode_FM02_S4.pdf"))
bartools::plotUmiFilterThresholds(timemachine_s4)
ggsave(file.path(plots_dir, "FateMap_umi_filter_thresholds_FM02_S4.pdf"))
```

### Load BARtab results

Read in BARtab results.

```{r}
bartab_s1 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp/counts/1-1uMPLX_S1.counts.tsv"))
bartab_s2 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp/counts/2-100nMPLX_S2.counts.tsv"))
bartab_s3 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp/counts/3-5nMT_S3.counts.tsv"))
bartab_s4 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp/counts/4-5nMT100nMP_S4.counts.tsv"))
```

Read in BARtab results without UMI error correction. 

```{r}
bartab_no_umi_s1 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp_no_umi/counts/1-1uMPLX_S1.counts.tsv"))
bartab_no_umi_s2 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp_no_umi/counts/2-100nMPLX_S2.counts.tsv"))
bartab_no_umi_s3 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp_no_umi/counts/3-5nMT_S3.counts.tsv"))
bartab_no_umi_s4 <- bartools::readBartabCounts(file.path(results_dir, "bartab_results_fm02_d8_50bp_no_umi/counts/4-5nMT100nMP_S4.counts.tsv"))
```


```{r}
bartools::plotBarcodesPerCell(bartab_s1)
ggsave(file.path(plots_dir, "BARtab_barcodes_per_cell_FM02_S1.pdf"))
bartools::plotUmiPerBarcode(bartab_s1)
ggsave(file.path(plots_dir, "BARtab_umi_per_barcode_FM02_S1.pdf"))
bartools::plotUmiFilterThresholds(bartab_s1)
ggsave(file.path(plots_dir, "BARtab_umi_filter_thresholds_FM02_S1.pdf"))

bartools::plotBarcodesPerCell(bartab_s2)
ggsave(file.path(plots_dir, "BARtab_barcodes_per_cell_FM02_S2.pdf"))
bartools::plotUmiPerBarcode(bartab_s2)
ggsave(file.path(plots_dir, "BARtab_umi_per_barcode_FM02_S2.pdf"))
bartools::plotUmiFilterThresholds(bartab_s2)
ggsave(file.path(plots_dir, "BARtab_umi_filter_thresholds_FM02_S2.pdf"))

bartools::plotBarcodesPerCell(bartab_s3)
ggsave(file.path(plots_dir, "BARtab_barcodes_per_cell_FM02_S3.pdf"))
bartools::plotUmiPerBarcode(bartab_s3)
ggsave(file.path(plots_dir, "BARtab_umi_per_barcode_FM02_S3.pdf"))
bartools::plotUmiFilterThresholds(bartab_s3)
ggsave(file.path(plots_dir, "BARtab_umi_filter_thresholds_FM02_S3.pdf"))

bartools::plotBarcodesPerCell(bartab_s4)
ggsave(file.path(plots_dir, "BARtab_barcodes_per_cell_FM02_S4.pdf"))
bartools::plotUmiPerBarcode(bartab_s4)
ggsave(file.path(plots_dir, "BARtab_umi_per_barcode_FM02_S4.pdf"))
bartools::plotUmiFilterThresholds(bartab_s4)
ggsave(file.path(plots_dir, "BARtab_umi_filter_thresholds_FM02_S4.pdf"))

bartools::plotBarcodesPerCell(bartab_no_umi_s1)
bartools::plotUmiPerBarcode(bartab_no_umi_s1)
bartools::plotUmiFilterThresholds(bartab_no_umi_s1)
```


## Compare FateMap vs. BARtab

### Number of cells annotated with barcodes

Across all conditions


BARtab 39636 cells, FateMap 39663 cells, whitelists 39884 cells

VennDiagram of cells annotated with barcodes

```{r}
bartab_cells <-
  rbind(
    mutate(bartab_s1, cellid_sample = paste0(cellid, "_S1")),
    mutate(bartab_s2, cellid_sample = paste0(cellid, "_S2")),
    mutate(bartab_s3, cellid_sample = paste0(cellid, "_S3")),
    mutate(bartab_s4, cellid_sample = paste0(cellid, "_S4"))
  )

timemachine_cells <-
  rbind(
    mutate(timemachine_s1, cellid_sample = paste0(cellid, "_S1")),
    mutate(timemachine_s2, cellid_sample = paste0(cellid, "_S2")),
    mutate(timemachine_s3, cellid_sample = paste0(cellid, "_S3")),
    mutate(timemachine_s4, cellid_sample = paste0(cellid, "_S4"))
  )

whitelist_cells <-
  c(
    paste0(whitelist_s1, "_S1"),
    paste0(whitelist_s2, "_S2"),
    paste0(whitelist_s3, "_S3"),
    paste0(whitelist_s4, "_S4")
  )
```

```{r}
length(unique(bartab_cells$cellid_sample))
length(unique(timemachine_cells$cellid_sample))
length(whitelist_cells)
```

```{r}
x <- list(
  BARtab = sort(unique(bartab_cells$cellid_sample)),
  FateMap = sort(unique(timemachine_cells$cellid_sample)),
  Whitelists = sort(whitelist_cells)
)

ggvenn(
  x,
  columns = c("BARtab", "FateMap", "Whitelists"),
  fill_color = c("#0073C2FF", "#EFC000FF", "#A83232"),
  stroke_size = 0.5,
  set_name_size = 5
)

ggsave(
  file.path(
    plots_dir,
    "venn_cells_with_barcodes_bartab_fatemap_whitelists_FM02.pdf"
  ),
  width = 7,
  height = 5
)
```

```{r}
ggvenn(
  x,
  columns = c("BARtab", "FateMap"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5,
  set_name_size = 5
)

ggsave(
  file.path(
    plots_dir,
    "venn_cells_with_barcodes_bartab_fatemap_FM02.pdf"
  ),
  width = 7,
  height = 5
)
```


### Number and percent of cells with 1 barcode annotated

Count number of cells annotated with 1 barcode at different UMI thresholds and the percentage of cells annotated with 1 barcode. 

By increasing the UMI threshold, you will get cleaner data and a higher percentage of cells with 1 barcode but at the same time overall potentially annotate less cells. 

Compare between BARtab, BARtab without umi correction and Timemachine for all 4 samples in FM02. 

```{r}
count_singlets_thresholds <-
  function(bc_objects, filter_range = 15) {
    singlet_res <- lapply(bc_objects, function (bc) {
      res_filtering <- list()
      # umi cutoff is increasing
      # to save compute time, use the filtered object to increase the umi threshold.
      bc_filtered <- bc
      for (i in seq(filter_range)) {
        bc_filtered <-
          bartools::filterBarcodes(bc_filtered, umiCountFilter = i, umiFractionFilter = 0)
        bc_singlets <- bc_filtered %>%
          group_by(cellid) %>%
          mutate(nLineages = length(cellid)) %>%
          filter(nLineages <= 1) %>%
          nrow()
        
        # perecent singlets of cells annotated with cutoff
        bc_perc_singlets <-
          bc_singlets / length(unique(bc_filtered$cellid))
        
        # percent singlets of all cells annotated
        bc_perc_singlets_total <-
          bc_singlets / length(unique(bc$cellid))
        
        res_filtering[[i]] <- c(bc_singlets, bc_perc_singlets, bc_perc_singlets_total)
      }
      return(res_filtering)
    })
    return(singlet_res)
  }
```

```{r}
singlet_counts_s1 <-
  count_singlets_thresholds(list(bartab_s1, bartab_no_umi_s1, timemachine_s1), 15)
singlet_counts_s2 <-
  count_singlets_thresholds(list(bartab_s2, bartab_no_umi_s2, timemachine_s2), 15)
singlet_counts_s3 <-
  count_singlets_thresholds(list(bartab_s3, bartab_no_umi_s3, timemachine_s3), 15)
singlet_counts_s4 <-
  count_singlets_thresholds(list(bartab_s4, bartab_no_umi_s4, timemachine_s4), 15)
```

```{r}
res_filtering_df_s1 <- cbind(data.frame(t(sapply(
  lapply(singlet_counts_s1[[1]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s1[[2]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s1[[3]], unlist), c
))))
colnames(res_filtering_df_s1) <-
  c(
    "bartab_cells",
    "bartab_cells_perc",
    "bartab_cells_perc_total",
    "bartab_no_umi_cells",
    "bartab_no_umi_cells_perc",
    "bartab_no_umi_cells_perc_total",
    "timemachine_cells",
    "timemachine_cells_perc",
    "timemachine_cells_perc_total"
  )
res_filtering_df_s1$umi_threshold <- seq(15)
res_filtering_df_s1$sample <- "S1"

res_filtering_df_s2 <- cbind(data.frame(t(sapply(
  lapply(singlet_counts_s2[[1]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s2[[2]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s2[[3]], unlist), c
))))
colnames(res_filtering_df_s2) <-
  c(
    "bartab_cells",
    "bartab_cells_perc",
    "bartab_cells_perc_total",
    "bartab_no_umi_cells",
    "bartab_no_umi_cells_perc",
    "bartab_no_umi_cells_perc_total",
    "timemachine_cells",
    "timemachine_cells_perc",
    "timemachine_cells_perc_total"
  )
res_filtering_df_s2$umi_threshold <- seq(15)
res_filtering_df_s2$sample <- "S2"

res_filtering_df_s3 <- cbind(data.frame(t(sapply(
  lapply(singlet_counts_s3[[1]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s3[[2]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s3[[3]], unlist), c
))))
colnames(res_filtering_df_s3) <-
  c(
    "bartab_cells",
    "bartab_cells_perc",
    "bartab_cells_perc_total",
    "bartab_no_umi_cells",
    "bartab_no_umi_cells_perc",
    "bartab_no_umi_cells_perc_total",
    "timemachine_cells",
    "timemachine_cells_perc",
    "timemachine_cells_perc_total"
  )
res_filtering_df_s3$umi_threshold <- seq(15)
res_filtering_df_s3$sample <- "S3"

res_filtering_df_s4 <- cbind(data.frame(t(sapply(
  lapply(singlet_counts_s4[[1]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s4[[2]], unlist), c
))), data.frame(t(sapply(
  lapply(singlet_counts_s4[[3]], unlist), c
))))
colnames(res_filtering_df_s4) <-
  c(
    "bartab_cells",
    "bartab_cells_perc",
    "bartab_cells_perc_total",
    "bartab_no_umi_cells",
    "bartab_no_umi_cells_perc",
    "bartab_no_umi_cells_perc_total",
    "timemachine_cells",
    "timemachine_cells_perc",
    "timemachine_cells_perc_total"
  )
res_filtering_df_s4$umi_threshold <- seq(15)
res_filtering_df_s4$sample <- "S4"
```

```{r}
combined_res_filtering_df <-
  rbind(
    res_filtering_df_s1,
    res_filtering_df_s2,
    res_filtering_df_s3,
    res_filtering_df_s4
  )
```

```{r}
head(combined_res_filtering_df)
```

Save intermediate results.

```{r}
write.csv(combined_res_filtering_df,
          file.path(results_dir, "fm02_umi_threshold_comp.csv"))
```

Plot number and percent of cells with a single barcode annotated. 

```{r}
combined_res_filtering_df %>%
  rename(
    `BARtab` = bartab_cells,
    `BARtab w/o UMI\nerror correction` = bartab_no_umi_cells,
    FateMap = timemachine_cells
  ) %>%
  pivot_longer(
    cols = c(`BARtab`, `BARtab w/o UMI\nerror correction`, `FateMap`),
    names_to = "method",
    values_to = "cells"
  ) %>%
  ggplot(aes(x = umi_threshold, y = cells, color = method)) +
  geom_point() +
  geom_line() +
  facet_wrap( ~ sample) +
  ylab("Cells with 1 barcode annotated") +
  xlab("UMI threshold") +
  theme_bw()

ggsave(
  file.path(plots_dir, "cells_1_barcode_FM02.pdf"),
  width = 7,
  height = 5
)
```

```{r}
combined_res_filtering_df %>%
  rename(
    `BARtab` = bartab_cells_perc,
    `BARtab w/o UMI\nerror correction` = bartab_no_umi_cells_perc,
    FateMap = timemachine_cells_perc
  ) %>%
  pivot_longer(
    cols = c(`BARtab`, `BARtab w/o UMI\nerror correction`, `FateMap`),
    names_to = "method_perc",
    values_to = "cells_perc"
  ) %>%
  ggplot(aes(x = umi_threshold, y = cells_perc * 100, color = method_perc)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ sample) +
  ylab("Cells with 1 barcode annotated (%)") +
  xlab("UMI threshold") +
  theme_bw()

ggsave(
  file.path(plots_dir, "perc_cells_1_barcode_FM02.pdf"),
  width = 7,
  height = 5
)
```

```{r}
combined_res_filtering_df %>%
  rename(
    `BARtab` = bartab_cells_perc_total,
    `BARtab w/o UMI\nerror correction` = bartab_no_umi_cells_perc_total,
    FateMap = timemachine_cells_perc_total
  ) %>%
  pivot_longer(
    cols = c(`BARtab`, `BARtab w/o UMI\nerror correction`, `FateMap`),
    names_to = "method_perc",
    values_to = "cells_perc"
  ) %>%
  ggplot(aes(x = umi_threshold, y = cells_perc * 100, color = method_perc)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ sample) +
  ylab("Cells with 1 barcode annotated (%)") +
  xlab("UMI threshold") +
  theme_bw()

ggsave(
  file.path(plots_dir, "perc_total_cells_1_barcode_FM02.pdf"),
  width = 7,
  height = 5
)
```


Only sample 1

```{r fig.height=3.5}
combined_res_filtering_df %>%
  filter(sample == "S1") %>%
  rename(
    `BARtab` = bartab_cells,
    `BARtab w/o UMI\nerror correction` = bartab_no_umi_cells,
    FateMap = timemachine_cells
  ) %>%
  pivot_longer(
    cols = c(`BARtab`, `BARtab w/o UMI\nerror correction`, `FateMap`),
    names_to = "method",
    values_to = "cells"
  ) %>%
  ggplot(aes(x = umi_threshold, y = cells, color = method)) +
  geom_point() +
  geom_line() +
  ylab("Cells with 1 barcode annotated") +
  xlab("UMI threshold") +
  theme_bw()

ggsave(
  file.path(plots_dir, "cells_1_barcode_FM02_S1.pdf"),
  width = 4.9,
  height = 3.5
)
```


```{r fig.height=3.5}
combined_res_filtering_df %>%
  filter(sample == "S1") %>%
  rename(
    `BARtab` = bartab_cells_perc,
    `BARtab w/o UMI\nerror correction` = bartab_no_umi_cells_perc,
    FateMap = timemachine_cells_perc
  ) %>%
  pivot_longer(
    cols = c(`BARtab`, `BARtab w/o UMI\nerror correction`, `FateMap`),
    names_to = "method_perc",
    values_to = "cells_perc"
  ) %>%
  ggplot(aes(x = umi_threshold, y = cells_perc * 100, color = method_perc)) +
  geom_point() +
  geom_line() +
  ylab("Cells with 1 barcode annotated (%)") +
  xlab("UMI threshold") +
  theme_bw()

ggsave(
  file.path(plots_dir, "perc_cells_1_barcode_FM02_S1.pdf"),
  width = 4.9,
  height = 3.5
)
```

```{r}
combined_res_filtering_df %>%
  filter(sample == "S1") %>%
  rename(
    `BARtab` = bartab_cells_perc_total,
    `BARtab w/o UMI\nerror correction` = bartab_no_umi_cells_perc_total,
    FateMap = timemachine_cells_perc_total
  ) %>%
  pivot_longer(
    cols = c(`BARtab`, `BARtab w/o UMI\nerror correction`, `FateMap`),
    names_to = "method_perc",
    values_to = "cells_perc"
  ) %>%
  ggplot(aes(x = umi_threshold, y = cells_perc * 100, color = method_perc)) +
  geom_point() +
  geom_line() +
  ylab("Cells with 1 barcode annotated (%)") +
  xlab("UMI threshold") +
  theme_bw()

ggsave(
  file.path(plots_dir, "perc_total_cells_1_barcode_FM02_S1.pdf"),
  width = 4.9,
  height = 3.5
)
```



#####################################################################################
#####################################################################################

### Comparison detected barcodes and clone size

Filtering barcodes based on UMI count

```{r}
bartab_s1_filtered <-
  bartools::filterBarcodes(bartab_s1,
                           umiCountFilter = 5,
                           umiFractionFilter = 0)
timemachine_s1_filtered <-
  bartools::filterBarcodes(timemachine_s1,
                           umiCountFilter = 15,
                           umiFractionFilter = 0)

bartab_s2_filtered <-
  bartools::filterBarcodes(bartab_s2,
                           umiCountFilter = 5,
                           umiFractionFilter = 0)
timemachine_s2_filtered <-
  bartools::filterBarcodes(timemachine_s2,
                           umiCountFilter = 15,
                           umiFractionFilter = 0)

bartab_s3_filtered <-
  bartools::filterBarcodes(bartab_s3,
                           umiCountFilter = 5,
                           umiFractionFilter = 0)
timemachine_s3_filtered <-
  bartools::filterBarcodes(timemachine_s3,
                           umiCountFilter = 15,
                           umiFractionFilter = 0)

bartab_s4_filtered <-
  bartools::filterBarcodes(bartab_s4,
                           umiCountFilter = 5,
                           umiFractionFilter = 0)
timemachine_s4_filtered <-
  bartools::filterBarcodes(timemachine_s4,
                           umiCountFilter = 15,
                           umiFractionFilter = 0)
```

Examplary QC plot for filtered data

```{r}
bartools::plotBarcodesPerCell(bartab_s1_filtered)
ggsave(
  file.path(plots_dir, "BARtab_barcodes_per_cell_FM02_S1_filtered.pdf"),
  width = 7,
  height = 5
)
bartools::plotBarcodesPerCell(timemachine_s1_filtered)
ggsave(
  file.path(plots_dir, "FateMap_barcodes_per_cell_FM02_S1_filtered.pdf"),
  width = 7,
  height = 5
)
```


Aggregate barcodes per cell and remove all cells with multiple barcodes detected (as it is done in FateMap publication).

```{r}
bartab_s1_agg <- bartools::aggregateBarcodes(bartab_s1_filtered)
timemachine_s1_agg <-
  bartools::aggregateBarcodes(timemachine_s1_filtered)

bartab_s1_singlets <- bartab_s1_agg %>%
  filter(!grepl(";", barcode))

timemachine_s1_singlets <- timemachine_s1_agg %>%
  filter(!grepl(";", barcode))

bartab_s2_agg <- bartools::aggregateBarcodes(bartab_s2_filtered)
timemachine_s2_agg <-
  bartools::aggregateBarcodes(timemachine_s2_filtered)

bartab_s2_singlets <- bartab_s2_agg %>%
  filter(!grepl(";", barcode))

timemachine_s2_singlets <- timemachine_s2_agg %>%
  filter(!grepl(";", barcode))

bartab_s3_agg <- bartools::aggregateBarcodes(bartab_s3_filtered)
timemachine_s3_agg <-
  bartools::aggregateBarcodes(timemachine_s3_filtered)

bartab_s3_singlets <- bartab_s3_agg %>%
  filter(!grepl(";", barcode))

timemachine_s3_singlets <- timemachine_s3_agg %>%
  filter(!grepl(";", barcode))

bartab_s4_agg <- bartools::aggregateBarcodes(bartab_s4_filtered)
timemachine_s4_agg <-
  bartools::aggregateBarcodes(timemachine_s4_filtered)

bartab_s4_singlets <- bartab_s4_agg %>%
  filter(!grepl(";", barcode))

timemachine_s4_singlets <- timemachine_s4_agg %>%
  filter(!grepl(";", barcode))
```


Venn diagrams of barcodes detected

```{r}
x <- list(
  FateMap_S1 = unique(timemachine_s1_singlets$barcode),
  BARtab_S1 = unique(bartab_s1_singlets$barcode),
  FateMap_S2 = unique(timemachine_s2_singlets$barcode),
  BARtab_S2 = unique(bartab_s2_singlets$barcode),
  FateMap_S3 = unique(timemachine_s3_singlets$barcode),
  BARtab_S3 = unique(bartab_s3_singlets$barcode),
  FateMap_S4 = unique(timemachine_s4_singlets$barcode),
  BARtab_S4 = unique(bartab_s4_singlets$barcode)
  )

ggvenn(
  x, columns = c("FateMap_S1", "BARtab_S1"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )

ggvenn(
  x, columns = c("FateMap_S2", "BARtab_S2"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )

ggvenn(
  x, columns = c("FateMap_S3", "BARtab_S3"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )

ggvenn(
  x, columns = c("FateMap_S4", "BARtab_S4"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )
```

Barcodes detected in at least 2 cells. 

```{r}
x_2c <- list(
  FateMap_S1 = unique(timemachine_s1_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  BARtab_S1 = unique(bartab_s1_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  FateMap_S2 = unique(timemachine_s2_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  BARtab_S2 = unique(bartab_s2_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  FateMap_S3 = unique(timemachine_s3_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  BARtab_S3 = unique(bartab_s3_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  FateMap_S4 = unique(timemachine_s4_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode)),
  BARtab_S4 = unique(bartab_s4_singlets %>% group_by(barcode) %>% summarise(count = n()) %>% filter(count > 1) %>% pull(barcode))
  )

ggvenn(
  x_2c, columns = c("FateMap_S1", "BARtab_S1"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )

ggvenn(
  x_2c, columns = c("FateMap_S2", "BARtab_S2"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )

ggvenn(
  x_2c, columns = c("FateMap_S3", "BARtab_S3"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )

ggvenn(
  x_2c, columns = c("FateMap_S4", "BARtab_S4"),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5, set_name_size = 5
  )
```


### Correlation clone size by #cells annotated

```{r}
s1_clone_size_merged <-
  merge(
    bartab_s1_singlets %>% group_by(barcode) %>% summarise(count = n()),
    timemachine_s1_singlets %>% group_by(barcode) %>% summarise(count = n()),
    by = "barcode",
    suffixes = c("_BARtab", "_FateMap"),
    all = T
  ) %>%
  mutate(sample = "S1")

s2_clone_size_merged <-
  merge(
    bartab_s2_singlets %>% group_by(barcode) %>% summarise(count = n()),
    timemachine_s2_singlets %>% group_by(barcode) %>% summarise(count = n()),
    by = "barcode",
    suffixes = c("_BARtab", "_FateMap"),
    all = T
  ) %>%
  mutate(sample = "S2")

s3_clone_size_merged <-
  merge(
    bartab_s3_singlets %>% group_by(barcode) %>% summarise(count = n()),
    timemachine_s3_singlets %>% group_by(barcode) %>% summarise(count = n()),
    by = "barcode",
    suffixes = c("_BARtab", "_FateMap"),
    all = T
  ) %>%
  mutate(sample = "S3")

s4_clone_size_merged <-
  merge(
    bartab_s4_singlets %>% group_by(barcode) %>% summarise(count = n()),
    timemachine_s4_singlets %>% group_by(barcode) %>% summarise(count = n()),
    by = "barcode",
    suffixes = c("_BARtab", "_FateMap"),
    all = T
  ) %>%
  mutate(sample = "S4")
```

```{r}
all_samples_clone_size <-
  rbind(
    s1_clone_size_merged,
    s2_clone_size_merged,
    s3_clone_size_merged,
    s4_clone_size_merged
  ) %>%
  mutate(
    count_BARtab = replace_na(count_BARtab, 0),
    count_FateMap = replace_na(count_FateMap, 0)
  )
```

```{r}
write.csv(
  all_samples_clone_size,
  file.path(results_dir, "fm02_barcode50_clone_sizes.csv")
)
```

Calculate correlation of clone sizes between FateMap and BARtab.

```{r}
sp_cor_ls <- c()
ps_cor_ls <- c()

for (sample in c("S1", "S2", "S3", "S4")) {
  bc_counts_merged <- all_samples_clone_size %>%
    filter(sample == !!sample)
  
  sp_c <-
    cor(bc_counts_merged$count_BARtab,
        bc_counts_merged$count_FateMap,
        method = "spearman")
  sp_cor_ls <- c(sp_cor_ls, sp_c)
  ps_c <-
    cor(bc_counts_merged$count_BARtab,
        bc_counts_merged$count_FateMap,
        method = "pearson")
  ps_cor_ls <- c(ps_cor_ls, ps_c)
}
```

```{r}
cor_results <-
  data.frame(list(
    "spearman" = sp_cor_ls,
    "pearson" = ps_cor_ls,
    "sample" = c("S1", "S2", "S3", "S4")
  ))

cor_results <- cor_results %>%
  pivot_longer(cols = c(spearman, pearson), names_to = "method")
```

```{r}
cor_labels <- cor_results %>%
  filter(method == "pearson") %>%
  mutate(label = paste0(sample, " (r=", round(value, 3), ")")) %>%
  pull(label)

names(cor_labels) <- cor_results %>%
  filter(method == "pearson") %>% pull(sample)
```



```{r}
all_samples_clone_size %>%
  ggplot(aes(x=count_BARtab, y=count_FateMap)) +
  geom_abline(slope=1, intercept = 0, linetype=2, alpha=0.4, color="blue") +
  geom_point(alpha=0.3, size=1) +
  coord_equal() +
  facet_wrap(~sample,
             labeller = labeller(sample = cor_labels)) +
  ylab("clone size FateMap") +
  xlab("clone size BARtab") +
  theme_bw()

ggsave(file.path(plots_dir, "clone_size_filtered_1bc_FM02.pdf"))
```

